Vagueness \& Rationality

### 2018-09-02 springer/current/e1.tex
Added separate analyses for V at each level of selection task for D and E.

30 participants were recruited. Response times from all trials were trimmed at 2.5 standard deviations for each subject, leading to the loss of 236 trials, 3.1\% of the data. The distribution of remaining response times was skewed with many long responses. These remaining response times were log-transformed which reduced this skew so that their distribution more closely approximated a normal distribution. Condition means for response times and proportion of borderline responses are given in Figure \ref{resultsC-exp-1}.

\begin{figure}[htbp]
\centering
\includegraphics[trim = 20mm 0mm 30mm 0mm, clip, width=.49\textwidth]{figures/Ce1-rtplot-1.pdf}
\includegraphics[trim = 20mm 0mm 25mm 0mm, clip, width=.49\textwidth]{figures/Ce1-blBarChart-1}
\caption{Experiment 1 results: mean response times by condition and item, and counts of borderline case responses by condition.}
\label{resultsC-exp-1}
\end{figure}

@@

However, given that the response time plot in the left panel of Figure \ref{resultsC-exp-1} shows that responses to 6:15:24 in the "crisp numeric" instructions condition were extremely fast relative to the "vague numeric" instructions to 6:15:24, the effects in the model of the full dataset could be driven by this difference. A clearer picture of the effects of interest might be obtained by removing the 6:15:24 level of Item from the data set, and fitting the model to this restricted data. Doing this did not affect the direction of the effects in the full dataset, but whereas the effects were significant in the full dataset, they were not significant in the restricted dataset. Full details of the analysis of the restricted dataset are available in the online materials.

A generalized linear mixed model \citet{jaeger2008categorical} was fit to the data for selection of the borderline response, with sum-coded vagueness, instruction format, (and their interaction), and item as fixed effects, and the same effects as slopes over participant, as well as per-participant intercepts as random effects. The distribution of responses over the nearest match square, the borderline square, and the furthest match square are given in Figure \ref{resultsC-exp-1}. Participants chose the borderline square on $16.6\%$ of trials overall.

\begin{description}
	\item [Hypothesis 4 (Response Type)] Participants were significantly more likely to choose the borderline option for vague instructions than for crisp instructions (21.9\% vs 11.3\%: $\beta=0.62$, $se=0.22$, $z=2.8$, $p=0.0059$). Participants were also significantly more likely to choose the borderline square when the instruction used the numerical format rather than the verbal format (30.1\% vs 3.0\%: $\beta=-3.35$, $se=0.23$, $z=-14.6$, $p<0.0001$). 
\end{description}

>>>>

\paragraph{\textbf{Response times}}

30 participants were recruited. Response times from all trials were trimmed at 2.5 standard deviations for each subject, leading to the loss of 236 trials, 3.1\% of the data. The distribution of remaining response times was skewed with many long responses. These remaining response times were log-transformed which reduced this skew so that their distribution more closely approximated a normal distribution. Condition means for response times are given in Figure \ref{resultsC-exp-1-RT}.

\begin{figure}[htbp]
\centering
\includegraphics[width=\textwidth]{figures/Ce1-rtplot-1.pdf}
\caption{Experiment 1 results: mean response times by condition and item}
\label{resultsC-exp-1-RT}
\end{figure}

@@

However, given that the response time plot in Figure \ref{resultsC-exp-1-RT} shows that responses to 6:15:24 in the "crisp numeric" instructions condition were extremely fast relative to the "vague numeric" instructions to 6:15:24, the effects in the model of the full dataset could be driven by this difference. A clearer picture of the effects of interest might be obtained by removing the 6:15:24 level of Item from the data set, and fitting the model to this restricted data. Doing this did not affect the direction of the effects in the full dataset, but whereas the effects were significant in the full dataset, they were not significant in the restricted dataset. Full details of the analysis of the restricted dataset are available in the online materials.

\paragraph{\textbf{Borderline cases}}

A generalized linear mixed model \citet{jaeger2008categorical} was fit to the data for the distribution of responses indicating the borderline response, with sum-coded vagueness, instruction format, (and their interaction), and item as fixed effects, and the same effects as slopes over participant, as well as per-participant intercepts as random effects. The distribution of responses over the nearest match square, the borderline square, and the furthest match square are given in Figure \ref{resultsC-exp-1-BL}. Participants chose the borderline square on $16.6\%$ of trials overall.

\begin{figure}[htbp]
\centering
\includegraphics[width=\textwidth]{figures/Ce1-blBarChart-1}
\caption{Experiment 1 results: counts of borderline case responses by condition.}
\label{resultsC-exp-1-BL}
\end{figure}

\begin{description}
	\item [Test of Hypothesis 4 (Response Type)] Participants were significantly more likely to choose the borderline option for vague instructions than for crisp instructions (21.9\% vs 11.3\%: $\beta=0.62$, $se=0.22$, $z=2.8$, $p=0.0059$). Participants were also significantly more likely to choose the borderline square when the instruction used the numerical format rather than the verbal format (30.1\% vs 3.0\%: $\beta=-3.35$, $se=0.23$, $z=-14.6$, $p<0.0001$). 
\end{description}

### 2018-09-02 springer/current/e2.tex
Added separate analyses for V at each level of selection task for D and E.

\begin{description} 
	\item [Test of Hypothesis 1 (Crisp/Vague RT)] Vague instructions resulted in faster responses than crisp instructions on average. However this difference was not significant in the full model ($\beta=-0.0057$, $se=0.0137$, $t=-0.42$, $p=0.678$). Using Levy's method \citep{Levy:MainEffectsInteractions} to test for main effects in the presence of higher-order interactions, by doing model comparison between a null model that included all interaction terms involving Vagueness but leaving out a term for the main effect of Vagueness, against a full model that differed only by including Vagueness as a main effect, showed that the full model was no better than the reduced model ($df=1$, $p=0.676$), consituting more evidence that Vagueness did not exert a significant main effect on response times. 
	\item [Test of Hypothesis 2 (Comparison/Matching RT)] comparison instructions resulted in faster responses than matching instructions, and the difference was significant ($\beta=0.1618$, $se=0.0255$, $t=6.34$, $p<0.001$).
	\item [Test of Hypothesis 3 (Interaction)] Although Vagueness did not exert a significant main effect, Vagueness did exert effects in interactions with some other variables: the interaction between Vagueness and Selection task  was significant ($\beta=0.1306$, $se=0.0205$, $t=6.38$, $p<0.001$), suggesting that Vagueness speeded RTs in the comparison condition but slowed them down in the matching task. 
\end{description}

>>>>

\begin{description} 
	\item [Test of Hypothesis 1 (Crisp/Vague RT)] Vague instructions resulted in faster responses than crisp instructions on average. However this difference was not significant in the full model ($\beta=-0.0057$, $se=0.0137$, $t=-0.42$, $p=0.678$). Using Levy's method \citep{Levy:MainEffectsInteractions} to test for main effects in the presence of higher-order interactions, by doing model comparison between a null model that included all interaction terms involving Vagueness but leaving out a term for the main effect of Vagueness, against a full model that differed only by including Vagueness as a main effect, showed that the full model was no better than the reduced model ($df=1$, $p=0.676$), consituting more evidence that Vagueness did not exert a significant main effect on response times. 
	\item [Test of Hypothesis 2 (Comparison/Matching RT)] Instructions involving comparison resulted in faster responses than instructions involving matching, and the difference was significant ($\beta=0.1618$, $se=0.0255$, $t=6.34$, $p<0.001$).
	\item [Test of Hypothesis 3 (Interaction)] The interaction between Vagueness and Selection task was significant ($\beta=0.1306$, $se=0.0205$, $t=6.38$, $p<0.001$), showing that vagueness exerted different effects on response times according to whether the instructions involved comparison or matching. When separate analyses were carried out testing for the effect of vagueness in comparison-only and in matching-only conditions, vague instructions led to significantly \emph{faster} responses than crisp instructions in the comparison-only conditions ($\beta=-0.071$, $se=0.020$, $t=-3.52$, $p<0.01$); and in the matching-only conditions vague instructions led to significantly \emph{slower} responses than crisp instructions ($\beta=0.062$, $se=0.021$, $t=2.91$, $p<0.01$).
\end{description}

### 2018-09-02 springer/current/e3.tex
Added separate analyses for V at each level of selection task for D and E.

\begin{figure}[htbp]
\centering
\includegraphics[width=.5\textwidth]{figures/Ee3-flow.jpg}
\caption{Experiment 3 example stimulus}
\label{Experiment3examplestimulus}
\end{figure}

\begin{table}
\centering
\caption{Experiment 3 instructions arranged by condition. The instructions given in the table started with ``Choose a square with \ldots"}
\label{instructionsE-exp-3}
\resizebox{\textwidth}{!}{\begin{tabular}{ccccl}
\hline\noalign{\smallskip}
Item & Target & Selection & Vagueness & Instruction \\ 
\noalign{\smallskip}\hline\noalign{\smallskip}
06:15:24 &   6 & Matching 	& Crisp 	& the same number of dots as the target 	    \\ 
06:15:24 &  10 & Matching	& Vague		& about the same number of dots as the target	\\
06:15:24 &  20 & Comparison & Crisp 	& fewer dots than the target             	    \\ 
06:15:24 &  20 & Comparison	& Vague		& far fewer dots than the target				\\
\noalign{\smallskip}\hline
\end{tabular}}
\end{table}

@@

\begin{description}
	\item [Test of Hypothesis 1 (Crisp/Vague RT)] Vague instructions resulted in faster responses than crisp instructions on average. However this difference was not significant in the full model ($\beta=-0.015$, $se=0.0097$, $t=-1.58$, $p=0.119$). Using Levy's method \citep{Levy:MainEffectsInteractions} to test for main effects in the presence of higher-order interactions, by doing model comparison between a null model that included all interaction terms involving Vagueness but leaving out a term for the main effect of Vagueness, against a full model that differed only by including Vagueness as a main effect, showed that the full model was no better than the reduced model ($df=1$, $p=0.118$), consituting more evidence that Vagueness did not exert a significant main effect on response times. 
	\item [Test of Hypothesis 2 (Comparison/Matching RT)] Comparison instructions resulted in significantly faster responses than matching instructions ($\beta=0.176$, $se=0.0168$, $t=10.51$, $p<0.001$).
	\item [Test of Hypothesis 3 (Interaction)] Although Vagueness did not exert a significant main effect, Vagueness did exert effects in its interaction with Selection task ($\beta=0.123$, $se=0.0166$, $t=7.42$, $p<0.001$), suggesting that Vagueness speeded response times in the comparison condition but slowed response times in the matching task. 
\end{description}

>>>>

\begin{figure}[htbp]
\centering
\includegraphics[width=.5\textwidth]{figures/Ee3-flow.jpg}
\caption{Experiment 3 example stimulus, showing the sequence of presentation, with the instruction screen at the top, followed by the presentation of the "target" array, followed by the stimulus that participants responded to.}
\label{Experiment3examplestimulus}
\end{figure}

\begin{table}
\centering
\caption{Experiment 3 instructions for the smallest array, arranged by condition. The instructions given in the table started with ``Choose a square with \ldots"}
\label{instructionsE-exp-3}
\resizebox{\textwidth}{!}{\begin{tabular}{ccccl}
\hline\noalign{\smallskip}
Item & Target & Selection & Vagueness & Instruction \\ 
\noalign{\smallskip}\hline\noalign{\smallskip}
06:15:24 &   6 & Matching 	& Crisp 	& the same number of dots as the target 	    \\ 
06:15:24 &  10 & Matching	& Vague		& about the same number of dots as the target	\\
06:15:24 &  20 & Comparison & Crisp 	& fewer dots than the target             	    \\ 
06:15:24 &  20 & Comparison	& Vague		& far fewer dots than the target				\\
\noalign{\smallskip}\hline
\end{tabular}}
\end{table}

@@

\begin{description}
	\item [Test of Hypothesis 1 (Crisp/Vague RT)] Vague instructions resulted in faster responses than crisp instructions on average. However this difference was not significant in the full model ($\beta=-0.015$, $se=0.0097$, $t=-1.58$, $p=0.119$). Using Levy's method \citep{Levy:MainEffectsInteractions} to test for main effects in the presence of higher-order interactions, by doing model comparison between a null model that included all interaction terms involving Vagueness but leaving out a term for the main effect of Vagueness, against a full model that differed only by including Vagueness as a main effect, showed that the full model was no better than the reduced model ($df=1$, $p=0.118$), consituting more evidence that Vagueness did not exert a significant main effect on response times. 
	\item [Test of Hypothesis 2 (Comparison/Matching RT)] Comparison instructions resulted in significantly faster responses than matching instructions ($\beta=0.176$, $se=0.0168$, $t=10.51$, $p<0.001$).
	\item [Test of Hypothesis 3 (Interaction)] The interaction between Vagueness and Selection task was significant ($\beta=0.123$, $se=0.0166$, $t=7.42$, $p<0.001$), showing that Vagueness had a different influence on response times when the instructions mandated comparison versus when the instructions mandated matching. When separate analyses were carried out testing for the effect of vagueness in comparison-only and in matching-only conditions, vague instructions led to significantly \emph{faster} responses than crisp instructions in the comparison-only conditions ($\beta=-0.077$, $se=0.018$, $t=-4.31$, $p<0.001$); and in the matching-only conditions vague instructions led to significantly \emph{slower} responses than crisp instructions ($\beta=0.047$, $se=0.013$, $t=3.71$, $p<0.001$).
\end{description}

### 2018-09-02 springer/current/general-discussion.tex
Added separate analyses for V at each level of selection task for D and E.

A comparison may clarify the logic of the situation. In recent years a number of studies, focussing on red wine, have suggested that alcohol, consumed in low doses, may have health benefits. An alternative explanation, however, asserts that it is not the alcohol in the wine that was beneficial, but antioxidants from grapes. If this alternative explanation is correct, then alcohol may not be healthy after all.\\[2ex]
%
% KvD Feb 2017. I've diminished the role of NLG in the coming paragraph.
Our findings suggest a re-think of the questions on which much research on the utility of vagueness rests. Years of research on the logic of vagueness -- giving rise to such techniques as Partial Logic \citet[e.g.,]{Fine}, Probabilistic Logic \citet{Edgington}, and Fuzzy Logic \citet{Zadeh} -- have primed the research community to expect that some special utility of vagueness is an important part of the answer, but our findings call this expectation into question.

>>>>

A comparison may clarify the logic of the situation. In recent years a number of studies, focussing on red wine, have suggested that alcohol, consumed in low doses, may have health benefits. An alternative explanation, however, asserts that it is not the alcohol in the wine that was beneficial, but antioxidants from grapes. If this alternative explanation is correct, then alcohol may not be healthy after all.\\[2ex]
%
% KvD Feb 2017. I've diminished the role of NLG in the coming paragraph.
Our findings suggest a re-think of the questions on which much research on the utility of vagueness rests. Years of research on the logic of vagueness -- giving rise to such techniques as Partial Logic \citep[e.g.,]{Fine}, Probabilistic Logic \citep{Edgington}, and Fuzzy Logic \citep{Zadeh} -- have primed the research community to expect that some special utility of vagueness is an important part of the answer, but our findings call this expectation into question.
